Numerical Stability of Detonations in White Dwarf Simulations

### 2018-11-23 papers/ignition/paper.tex
Update discussion of suppressed burning limiter

\citet{kushnir:2013} observe an important possible failure mode
for reacting hydrodynamics simulations. Let us define $\tau_e = e / \dot{e}$
as the nuclear energy injection timescale, and $\tau_s = \Delta x / c_s$
as the sound-crossing time in a zone (where $\Delta x$ is the grid
resolution and $c_s$ is the speed of sound).
When the sound-crossing time is too long, energy is built up in
a zone faster than it can be advected away by pressure waves.
This is of course a problem inherent only to numerically discretized
systems as the underlying fluid equations are continuous.
This can lead to a numerically seeded detonation caused by the
temperature building up too quickly in the zone; the detonation
may be spurious in this case. If $\tau_s \ll \tau_e$, we can be
confident that a numerically seeded detonation has not occurred.
In practice, we will quantify this as:
\begin{equation}
  \tau_s \leq f_{s}\, \tau_e \label{eq:burning_limiter_2}
\end{equation}
and require that $f_{s}$ is sufficiently smaller than one.
\citet{kushnir:2013} state that $f_{s} = 0.1$ is a sufficient
criterion. \citet{kushnir:2013} enforced this criterion on
their simulations by artificially limiting the magnitude of the energy
release after a burn.

To understand this choice, consider the outcome of our baseline case.
When the two WDs collide, there is a prompt detonation
at the contact point at the center of the domain. This detonation occurs
when the density is relatively low because the WDs have not had much
time to collide, and so the nucleosynthetic yield of nickel is low. If
this is the case, the colliding WDs are not a good candidate for
normal SNe Ia. However, if this detonation is delayed, material builds
up at a stalled shock in the center, and the eventual detonation occurs
on the outer edge of the stalled shock. There will then be much more
high density material to process into nickel, and the resulting
explosion creates an amount of nickel potentially sufficient to explain
normal SNe Ia. The discussion thus centers on whether the prompt detonation
is physical. \citeauthor{kushnir:2013} argues that it is not, on the basis
that $\tau_s > \tau_e$ when this detonation occurs. We find the same thing:
for the baseline case, the sound-crossing time is about five times longer
than the energy injection time. This means that an ignition is guaranteed
to occur for numerical reasons. \citeauthor{kushnir:2013} thus argue that
the suppressed burn is justifiable, as it suppresses this prompt detonation.

This argument is flawed because a physical detonation may \textit{also}
occur this way. For example, consider a region of WD material at uniformly
high temperature, say $5 \times 10^9\ \text{K}$, with an arbitrarily large size,
say a cube with side length 100 km. This region will very likely ignite,
even if it is surrounded by much cooler material. By the time the material on
the edges can advect heat away, the material in the center will have long since
started burning carbon, as the sound crossing time scale is sufficiently large
compared to the energy injection time scale. This is true regardless of whether
the size of this cube corresponds to the spatial resolution in a simulation.
Suppression of the burn in this case is unphysical: if we have a zone matching
these characteristics, the zone should detonate.

When the resolution is low enough, there is a floor on the size of a hotspot,
possibly making such a detonation more likely. This is an unavoidable consequence
of the low resolution; yet, it may be the correct result of the simulation that
was performed. That is, even if large hotspots are unphysical because in reality
the temperature distribution would be smoother, if such a large hotspot \textit{were}
to develop (which is the implicit assumption of a low resolution simulation), then
it would likely detonate. If the results do not match what occurs at higher
resolution, then the simulation is not converged and the results are not reliable.
However, it may also be the case that a higher resolution simulation will yield
similar results, for example because even at the higher resolution, the physical
size of the hotspot stays the same. Additionally, one might think that simply
adding adaptive mesh refinement in the burning region could help, because then
$\Delta x$ decreases, so the ratio of $\tau_s$ to $\tau_e$ should decrease.
We developed an AMR criterion to do just this: we tag on regions suggested by
the burning stability criterion, \autoref{eq:burning_limiter_2}. But if a low
resolution simulation develops a very large hotspot that violates the stability
criterion, adding refinement inside this region will likely not help, because the
energy trapped in the center of the region may not be able to escape before ignition
occurs. So for the simulations with resolution inside the stars that was coarser than
10 km, even adding a significant amount of refinement (say, refining by a factor of
64 or 128) did not change the propagation of a detonation.

For the above reasons, an appeal to the numerical instability criterion alone is
insufficient to understand whether a given ignition is real. We find that \textit{all}
of the detonations we have seen at any resolution violate the stability criterion,
even for the finest resolution simulation we performed (150 m, inside the burning region).
This is true even if the prompt detonation is suppressed:
the delayed detonation also is numerically unstable. For comparison purposes, we
developed an option for our burner to use a ``suppressed'' burning mode. In a
suppressed burn, we limit the changes to the state so that \autoref{eq:burning_limiter_2}
is always satisfied\footnote{To achieve this we directly multiply the
right-hand-side vector in the integration by a factor $F$
for all variables, where $F$ is the multiplicative factor needed to
be applied to $\dot{e}$ such that the equality in \autoref{eq:burning_limiter_2}
holds. (If the inequality is already satisfied, then the integration
vector is not modified.) We fix $\tau_s$ to be the value of the sound
crossing time at the beginning of the burn (that is, we do not
update it as the sound speed changes) and we fix the energy $e$
that goes into the estimate for $\tau_e$ to be the value of the
internal energy of the zone at the beginning of the burn.}. When we
evolve the baseline case with this suppressed burner, the prompt detonation
is indeed suppressed. A later detonation does occur, but this delayed
detonation also triggers the suppression. The result is that the detonation
does not get to very high temperatures, and the result is again a ``failed''
supernova that does not generate nearly enough nickel. Thus we do not
reproduce the claim of \citeauthor{kushnir:2013} that the suppressed
burning mode suppresses only the prompt detonation, but allows the
delayed detonation to occur. We could find some way to turn off the
suppression and allow the delayed detonation, but without some independent
confirmation that the delayed detonation is physical (while the prompt
detonation is not) this would be begging the question.

The upshot of this discussion is that the only way to resolve the
question is with higher resolution simulations. The semi-analytic
criterion provided by \cite{garg:2017}, and the numerical calculations
of \cite{seitenzahl:2009}, give critical radii for spontaneous detonation.
For the densities we are interested in ($\sim 10^7\ \text{g\,/\,cm}^3$),
assuming equal C/O material, the critical radii are in the range 1-10 km.
So this suggests that our simulation needs to do significantly better than
this if we are likely to make headway on this issue. Detonations observed
at higher resolution may be physically plausible because the critical
temperature gradient can be resolved. The high resolution needs to be
achieved with some combination of high resolution of the collision process
itself, and AMR based on the stability criterion as described
above, where we tag for refinement all zones above some $f_{s}$, the
ratio of the sound crossing time to the energy injection time. We
cannot solely rely on a low resolution run with AMR based on the
burning stability criterion for the reason described above, that a
detonation may be spuriously locked in by the low resolution regardless
of how much resolution we can practicably add to it.

>>>>

\citet{kushnir:2013} observe an important possible failure mode
for reacting hydrodynamics simulations. Let us define $\tau_e = e / \dot{e}$
as the nuclear energy injection timescale, and $\tau_s = \Delta x / c_s$
as the sound-crossing time in a zone (where $\Delta x$ is the grid
resolution and $c_s$ is the speed of sound). When the sound-crossing
time is too long, energy is built up in a zone faster than it can be
advected away by pressure waves. This is of course a problem inherent
only to numerically discretized systems as the underlying fluid equations
are continuous. This can lead to a numerically seeded detonation
caused by the temperature building up too quickly in the zone; the
detonation may be spurious in this case. If $\tau_s \ll \tau_e$,
we can be confident that a numerically seeded detonation has not
occurred. In practice, we will quantify this requirement as:
\begin{equation}
  \tau_s \leq f_{s}\, \tau_e \label{eq:burning_limiter}
\end{equation}
and require that $f_{s}$ is sufficiently smaller than one.
\citet{kushnir:2013} state that $f_{s} = 0.1$ is a sufficient
criterion for avoiding premature ignitions. \citeauthor{kushnir:2013}
enforced this criterion on their simulations by artificially limiting
the magnitude of the energy release after a burn, and claimed that
this is resulted in more accurate WD collision simulations.

We find that for our test problem (and also the WD collisions we have simulated)
we do observe $\tau_s > \tau_e$; typically the ratio is a factor of 2--5 at
low resolution. This means that an ignition is very likely to occur for numerical
reasons, regardless of whether it would occur for physical reasons. We thus
investigate whether limiting the energy release of the burn (we will term this
``suppressing'' the burn) is a useful technique for avoiding the prompt detonation.

Before we examine the results, consider a flaw in the reasoning behind the limiter:
a physical detonation may \textit{also} occur with the property that, in the detonating
zone, $\tau_s > \tau_e$. For example, consider a region of WD material at uniformly
high temperature, say $5 \times 10^9\ \text{K}$, with an arbitrarily large size,
say a cube with side length 100 km. This region will very likely ignite,
even if it is surrounded by much cooler material. By the time the material on
the edges can advect heat away, the material in the center will have long since
started burning carbon, as the sound crossing time scale is sufficiently large
compared to the energy injection time scale. This is true regardless of whether
the size of this cube corresponds to the spatial resolution in a simulation.
Suppression of the burn in this case is unphysical: if we have a zone matching
these characteristics, the zone should detonate.

When the resolution is low enough, there is a floor on the size of a hotspot,
possibly making such a detonation more likely. This is an unavoidable consequence
of the low resolution; yet, it may be the correct result of the simulation that
was performed. That is, even if large hotspots are unphysical because in reality
the temperature distribution would be smoother, if such a large hotspot \textit{were}
to develop (which is the implicit assumption of a low resolution simulation), then
it would likely detonate. If the results do not match what occurs at higher
resolution, then the simulation is not converged and the results are not reliable.
However, it may also be the case that a higher resolution simulation will yield
similar results, for example because even at the higher resolution, the physical
size of the hotspot stays the same. For this reason, an appeal to the numerical
instability criterion alone is insufficient to understand whether a given ignition
is real.

\begin{figure}[ht]
  \centering
  \includegraphics[scale=0.30]{{{plots/amr_ignition_suppressed}}}
  \caption{Similar to \autoref{fig:self-heat-distance}, but for simulations with the
           suppressed burning limiter applied (\autoref{eq:burning_limiter}). Only
           simulations without AMR are shown.
           \label{fig:suppressed-distance}}
\end{figure}

\autoref{fig:suppressed-distance} shows the results we obtain for our implementation of
a ``suppressed'' burning mode. In a suppressed burn, we limit the changes to the state
so that \autoref{eq:burning_limiter} is always satisfied. This is done by rescaling the
energy release and species changes from a burn by a common factor such that the equality
in \autoref{eq:burning_limiter} is satisfied. (If the inequality is already satisfied,
then the integration vector is not modified.) We find that the suppressed burn
generally does not yield correct results for low resolutions. The 64 km resolution
simulation happens to yield approximately the correct ignition distance, but it does
not occur at the right time, and in any case the incorrectness of the results at neighboring
resolutions suggests that this is not a robust finding. Because of both the theoretical reasons
discussed above, and this empirical finding, we do not believe that the suppressed burning
limiter should be applied in production simulations.
